Development Of Parallel Matrix Operations

### 2011-02-15 paper/paper.tex
Add some explination's to the results

A Parallel Matrix Manipulation Library (PMML) is presented for use on sequential and symmetric multiprocessing (SMP) systems. PMML is written using the Parallel Java library developed by Professor Alan Kaminsky at the Rochester Institute of Technology. PMML provides methods for the following operations: matrix-vector multiplication, matrix-matrix multiplication, and matrix inversion and solution to the linear system $A x=b$. An implementation of the Strassen algorithm for efficient matrix multiplication is provided for use sequentially and on SMP systems. Gaussian elimination with partial-pivoting is used to find the row-reduced form of the matrix $A$ and backsubstitution is used to solve the linear system. Algorithms for sequential solution and parallel solution on SMP systems are provided. Performance metrics are presented to compare the {\em speedup} and {\em sizeup} gained in the parallel over the sequential versions.
\end{abstract}

@@

\begin{verbatim}
MatrixInt A = ...
VectorInt B = ...
Matrix C = MatrixOp.mult(A,B)
Matrix C = MatrixParallelOp.mult(A,B)

\end{verbatim}

@@

In the PMML, the following code:
\begin{verbatim}
Matrix C = MatrixOp.mult(A,B);
Matrix C = MatrixParallelOp.mult(A,B);
\end{verbatim}
returns the product $C=AB$, using the sequential and parallel algorithms respectively. Note that the operation is valid only if the number of columns of $A$ equals the number of rows of $B$, otherwise an Exception is thrown.

@@

@@

To solve a linear system of equations of the form 
\begin{equation}
$Ax=b$, 
\end{equation}
where $A$ is an $m\times n$ matrix, $b$ is an $m\times 1$ vector, and $x$ is the $m\times 1$ solution vector, first the matrix must be inverted. For most matrices, this presents a problem, as the inverse of a matrix is generally not easy to compute and may not even exist. Gaussian elimination is the process of performing a series of elementary row operations to the matrix $A$ to reduce it to a form, known as row-echelon form, that can be easily inverted. The matrix 
\begin{equation} 
\left[ \begin{array}{ccc|c}
a_{1,1} & a_{1,2} & a_{1,3} & b_1 \\
0 & a_{2,2} & a_{2,2} & b_2 \\
0 & 0 & a_{3,3} & b_3
\end{array} \right],
\end{equation}
where $a_{i,j}$ are the elements of $A$, and $b_i$ are the elements of $b$, is an example of the system $Ax=b$ in row-echelon form. Matrices in row-echelon form have the nice characteristic that all of the leading coefficients of each row are in a column whose number is less than the column of the leading coefficients in the rows below it and greater than those in the rows above it. In the process of Gaussian elimination with partial-pivoting, this is achieved by moving the row with the greatest leading coefficient above those with lesser coefficients, and then reducing the lower rows. In the case of square matrices whose rows are linearly independent, this reduction results in an upper-triangular form. Once the matrix is in this form, the inversion process is equivalent to back substituting the solution from the bottom of the matrix all the way to the top. Parallelization of this method consists of dividing up the task of reducing the rows of the matrix below the pivot point among threads.

@@

To test each of these methods, in the case of providing a matrix, a random matrix of size $n^2$ was generated by:
\begin{verbatim}
MatrixDouble A = MatrixDouble.random(1000, 1000);
\end{verbatim}
and in the case of a vector, a random vector of size $n$ was generated by:
\begin{verbatim}
VectorDouble b = VectorDouble.random(1000);
\end{verbatim}
Over 4 runs, the time required to perform the operation was averaged.

@@

\subsubsection{Sequential and Parallel Implementation}
\begin{center}
\begin{tabular}{|r|r|r|r|r|r|r|}\hline
{\em n} & {\em K} & {\em T (msec)} & {\em Speedup} & {\em Eff} & {\em EDSF} \\\hline
1000    & seq     &   32609        &   xxx         & xxx       & xxx        \\\hline
1000    & 1       &   53546        & 0.609        & 0.609    & xxx        \\\hline
1000    & 2       &   15812        & 2.062        & 1.031    & -0.409     \\\hline
1000    & 4       &    8101        & 4.025        & 1.006    & -0.132     \\\hline
1000    & 8       &    4703        & 6.934        & 0.991    & -0.064     \\\hline
1000    & 8       &    4210        & 7.746        & 0.968    & -0.053     \\\hline
500     & seq     &    4797        &   xxx         & xxx       & xxx        \\\hline
500     & 1       &    2046        & 2.345         & 2.345     & xxx        \\\hline
500     & 2       &    2632        & 1.823         & 0.911      & 1.573      \\\hline
500     & 4       &    1411        & 3.400         & 0.850     & 0.586      \\\hline
500     & 7       &     953        & 5.034         & 0.719     & 0.377      \\\hline
500     & 8       &     789        & 6.080         & 0.760     & 0.298      \\\hline
100     & seq     &      71        &   xxx         & xxx       & xxx        \\\hline
100     & 1       &      62        & 1.145        & 1.145    & xxx        \\\hline
100     & 2       &      42        & 1.690        & 0.845    & 0.355     \\\hline
100     & 4       &      36        & 1.972        & 0.493    & 0.441     \\\hline
100     & 7       &      35        & 2.029        & 0.290    & 0.492     \\\hline
100     & 8       &      35        & 2.029        & 0.254    & 0.502     \\\hline
\end{tabular}
\end{center}

\begin{center}
\includegraphics[scale=0.5]{images/parallel/run-vs-proc.png}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel/speedup-vs-proc.png}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel/edsf-vs-proc.png}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel/eff-vs-proc.png}
\end{center}

\subsubsection{Sequential Strassen and Parallel Strassen Implementation}
\begin{center}
\begin{tabular}{|r|r|r|r|r|r|r|}\hline
{\em n} & {\em K} & {\em T (msec)} & {\em Speedup} & {\em Eff} & {\em EDSF} \\\hline
1000    & seq     &   12296           &   xxx         & xxx       & xxx        \\\hline
1000    & 1       &   12105           &   1.016       & 1.016     & xxx        \\\hline
1000    & 2       &   7139            &   1.722       & 0.861     & 0.180      \\\hline
1000    & 4       &   4184            &   2.939       & 0.735     & 0.128      \\\hline
1000    & 8       &   2467            &   4.984       & 0.712     & 0.071      \\\hline
1000    & 8       &   2372            &   5.184       & 0.648     & 0.081      \\\hline
500     & seq     &   1792            &   xxx         & xxx       & xxx        \\\hline
500     & 1       &   1787            &   1.003       & 1.003     & xxx        \\\hline
500     & 2       &   1112            &   1.612       & 0.806     & 0.245      \\\hline
500     & 4       &   713             &   2.513       & 0.628     & 0.199      \\\hline
500     & 7       &   505             &   3.549       & 0.507     & 0.163      \\\hline
500     & 8       &   502             &   3.570       & 0.446     & 0.178      \\\hline
100     & seq     &   71              &   xxx         & xxx       & xxx        \\\hline
100     & 1       &   37              &   1.919       & 1.919     & xxx        \\\hline
100     & 2       &   69              &   1.029       & 0.514     & 2.730      \\\hline
100     & 4       &   70              &   1.014       & 0.254     & 2.189      \\\hline
100     & 7       &   68              &   1.044       & 0.149     & 1.977      \\\hline
100     & 8       &   68              &   1.044       & 0.131     & 1.958      \\\hline
\end{tabular}
\end{center}

\begin{center}
\includegraphics[scale=0.5]{images/parallel-stras/running-vs-proc.png}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel-stras/speedup-vs-proc.png}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel-stras/edsf-vs-proc.png}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel-stras/eff-vs-proc.png}
\end{center}

>>>>

A Parallel Matrix Manipulation Library (PMML) is presented for use on
sequential and symmetric multiprocessing (SMP) systems. PMML is written using the Parallel Java (PJ) library developed by Professor Alan Kaminsky at the Rochester Institute of Technology. PMML provides methods for the following operations: matrix-vector multiplication, matrix-matrix multiplication, and matrix inversion and solution to the linear system $A x=b$. An implementation of the Strassen algorithm for efficient matrix multiplication is provided for use sequentially and on SMP systems. Gaussian elimination with partial-pivoting is used to find the row-reduced form of the matrix $A$ and backsubstitution is used to solve the linear system. Algorithms for sequential solution and parallel solution on SMP systems are provided. Performance metrics are presented to compare the {\em speedup} and {\em sizeup} gained in the parallel over the sequential versions.
\end{abstract}

@@

@@

In the PMML, the following code:
\begin{verbatim}
MatrixInt A = ...
MatrixInt B = ...
Matrix C = MatrixOp.mult(A,B);
Matrix C = MatrixParallelOp.mult(A,B);
\end{verbatim}
returns the product $C=AB$, using the sequential and parallel algorithms respectively. Note that the operation is valid only if the number of columns of $A$ equals the number of rows of $B$, otherwise an Exception is thrown.

@@

{\tt MatrixInt A = ...\\
VectorInt v = ... \\
Matrix C = MatrixOp.mult(A,v) \\
Matrix C = MatrixParallelOp.mult(A,v) }

returns the product $C=A\cdot v$, using the sequential and parallel versions of the
matrix vector multiplication algorithm. Note the use of Java's polymorphism,
allowing a user to multiply a vector \& matrix, as
well as a matrix \& matrix using the same command.

@@

To solve a linear system of equations of the form 
\begin{equation}
Ax=b, 
\end{equation}
where $A$ is an $m\times n$ matrix, $b$ is an $m\times 1$ vector, and $x$ is the $m\times 1$ solution vector, first the matrix must be inverted. For most matrices, this presents a problem, as the inverse of a matrix is generally not easy to compute and may not even exist. Gaussian elimination is the process of performing a series of elementary row operations to the matrix $A$ to reduce it to a form, known as row-echelon form, that can be easily inverted. The matrix 
\begin{equation} 
\left[ \begin{array}{ccc|c}
a_{1,1} & a_{1,2} & a_{1,3} & b_1 \\
0 & a_{2,2} & a_{2,2} & b_2 \\
0 & 0 & a_{3,3} & b_3
\end{array} \right],
\end{equation}
where $a_{i,j}$ are the elements of $A$, and $b_i$ are the elements of $b$, is an example of the system $Ax=b$ in row-echelon form. Matrices in row-echelon form have the nice characteristic that all of the leading coefficients of each row are in a column whose number is less than the column of the leading coefficients in the rows below it and greater than those in the rows above it. In the process of Gaussian elimination with partial-pivoting, this is achieved by moving the row with the greatest leading coefficient above those with lesser coefficients, and then reducing the lower rows. In the case of square matrices whose rows are linearly independent, this reduction results in an upper-triangular form. Once the matrix is in this form, the inversion process is equivalent to back substituting the solution from the bottom of the matrix all the way to the top. Parallelization of this method consists of dividing up the task of reducing the rows of the matrix below the pivot point among threads.

@@

To test each of these methods, in the case of providing a matrix, a random matrix of size $n^2$ was generated by:

{\tt MatrixDouble A = MatrixDouble.random(1000, 1000); }

and in the case of a vector, a random vector of size $n$ was generated by:

{\tt VectorDouble b = VectorDouble.random(1000);}

Over 4 runs, the time required to perform the operation was averaged.
This technique was chosen in an attempt to force the JVM's Just in Time
Compiler to activate and produce machine code for the methods
being tested. This gives an average of the initial non-JIT timings, mixed
with the eventual JIT-enabled timings.

@@

\subsubsection{Sequential and Parallel Implementation}

\begin{center}
\begin{tabular}{|r|r|r|r|r|r|r|}\hline
{\em n} & {\em K} & {\em T (msec)} & {\em Speedup} & {\em Eff} & {\em EDSF} \\\hline
1000    & seq     &   32609        &   xxx         & xxx       & xxx        \\\hline
1000    & 1       &   53546        & 0.609        & 0.609    & xxx        \\\hline
1000    & 2       &   15812        & 2.062        & 1.031    & -0.409     \\\hline
1000    & 4       &    8101        & 4.025        & 1.006    & -0.132     \\\hline
1000    & 8       &    4703        & 6.934        & 0.991    & -0.064     \\\hline
1000    & 8       &    4210        & 7.746        & 0.968    & -0.053     \\\hline
500     & seq     &    4797        &   xxx         & xxx       & xxx        \\\hline
500     & 1       &    2046        & 2.345         & 2.345     & xxx        \\\hline
500     & 2       &    2632        & 1.823         & 0.911      & 1.573      \\\hline
500     & 4       &    1411        & 3.400         & 0.850     & 0.586      \\\hline
500     & 7       &     953        & 5.034         & 0.719     & 0.377      \\\hline
500     & 8       &     789        & 6.080         & 0.760     & 0.298      \\\hline
100     & seq     &      71        &   xxx         & xxx       & xxx        \\\hline
100     & 1       &      62        & 1.145        & 1.145    & xxx        \\\hline
100     & 2       &      42        & 1.690        & 0.845    & 0.355     \\\hline
100     & 4       &      36        & 1.972        & 0.493    & 0.441     \\\hline
100     & 7       &      35        & 2.029        & 0.290    & 0.492     \\\hline
100     & 8       &      35        & 2.029        & 0.254    & 0.502     \\\hline
\end{tabular}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel/run-vs-proc.png}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel/speedup-vs-proc.png}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel/edsf-vs-proc.png}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel/eff-vs-proc.png}
\end{center}
The results report strong parallel speedup especially as our matrix 
size increases. The speedup for the $1000 X 1000$ matrix multiplication reports
near linear speedup, and a efficiency of $1.0$ exactly. The particularly poor performance
for small matrices can potentially be traced back to the slight over head
that comes with starting and executing a Parallel Java team. It is possible that the time taken during
the initial dispatch of the threads in the parallel implementation is greater
than any possible speed up achieved by using them for small matrices.

\subsubsection{Sequential \& Parallel Strassen Implementation}

\begin{center}
\begin{tabular}{|r|r|r|r|r|r|r|}\hline
{\em n} & {\em K} & {\em T (msec)} & {\em Speedup} & {\em Eff} & {\em EDSF} \\\hline
1000    & seq     &   12296           &   xxx         & xxx       & xxx        \\\hline
1000    & 1       &   12105           &   1.016       & 1.016     & xxx        \\\hline
1000    & 2       &   7139            &   1.722       & 0.861     & 0.180      \\\hline
1000    & 4       &   4184            &   2.939       & 0.735     & 0.128      \\\hline
1000    & 8       &   2467            &   4.984       & 0.712     & 0.071      \\\hline
1000    & 8       &   2372            &   5.184       & 0.648     & 0.081      \\\hline
500     & seq     &   1792            &   xxx         & xxx       & xxx        \\\hline
500     & 1       &   1787            &   1.003       & 1.003     & xxx        \\\hline
500     & 2       &   1112            &   1.612       & 0.806     & 0.245      \\\hline
500     & 4       &   713             &   2.513       & 0.628     & 0.199      \\\hline
500     & 7       &   505             &   3.549       & 0.507     & 0.163      \\\hline
500     & 8       &   502             &   3.570       & 0.446     & 0.178      \\\hline
100     & seq     &   71              &   xxx         & xxx       & xxx        \\\hline
100     & 1       &   37              &   1.919       & 1.919     & xxx        \\\hline
100     & 2       &   69              &   1.029       & 0.514     & 2.730      \\\hline
100     & 4       &   70              &   1.014       & 0.254     & 2.189      \\\hline
100     & 7       &   68              &   1.044       & 0.149     & 1.977      \\\hline
100     & 8       &   68              &   1.044       & 0.131     & 1.958      \\\hline
\end{tabular}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel-stras/running-vs-proc.png}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel-stras/speedup-vs-proc.png}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel-stras/edsf-vs-proc.png}
\end{center}
\begin{center}
\includegraphics[scale=0.5]{images/parallel-stras/eff-vs-proc.png}
\end{center}
Overall the results for the sequential Strassen and Parallel-Strassen
algorithms reported significantly faster run-times when compared to the
sequential and parallel regular algorithms. However the speedup
compared to the other matrix multiplication results is not quite as
impressive. Their are many factors which contribute to this result.
First of all, for matrices smaller than size of 128 by 128 both of the
Strassen implementations default back to the regular sequential matrix
multiplication. During testing it was discovered that the regular sequential
algorithm is faster for this special case. This is why the results show no
speedup for 100 by 100 matrix. It should also be noted that in our parallel
Strassen implementation the maximum number of steps which can parallelized is
seven. As shown in the results above, when moving to eight parallel threads
very little speedup is achieved. The Strassen algorithms are also specially
designed to operate on matrices which are of a size which is a power of two.
In order to take advantage of this for ever matrix, if the matrix is not a
power of two the matrix is padded with zero's and then unpadded after the
algorithm is complete. Since none of the matrices tested were of a size which
was a power of two, the results shown are examples of the worst possible
running time for the sequential and parallel Strassen algorithms.
